---
title: "Gov 50 Final Project"
author: "Sebastian Juarez"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction
Do states that emit more carbon dioxide have a higher gross national income per capita? In my project, I will investigate the extent to which carbon dioxide output affects gross national income per capita. I hypothesize that higher carbon dioxide output will correlate with a higher gross national income per capita. When carbon dioxide production increases, I expect a higher number of goods and services created in and exported from the country, which boosts the economy and typically raises wages. This question is important because it can help inform whether or not sustainable development is a norm or even if it is possible. If countries have been able to reach their level of economic prosperity due to increased emissions of carbon, it can help answer questions of how to move forward with development in the future. It could also inform whether we should hold lesser developed countries up to the same standards of needing to mitigate carbon emissions to the same extent as developed countries, given that emissions are correlated with raising incomes.

My sample consists of data from countries across the globe with available data for that year, with data collected every year from 1990 to 2021. My unit of analysis is a country. If I observe an increase in gross national income where there are greater carbon dioxide production levels, this would support my hypothesis. If I observe a lower gross national income per capita in years where the level of carbon dioxide production has increased, this would prove against my hypothesis. If I find a strong, positive correlation between the gross national income per capita and the carbon dioxide output within countries, this would support my hypothesis.

## Data
My data is taken from countries all over the world with available data. Spanning the years 1990 to 2021, the data set includes valuable information on the Human Development Index (HDI), Life Expectancy, Gross National Income per capita (GNI), and CO2 Production. The data is taken from publicly available information about each country from the relevant agencies. The values are summarized in a table. I combined two data sets, one for GNI and one for CO2, for my study.

# Variables
This study uses a **cross-sectional observational** research design because the data was collected around the same time across the sample populations. The data points were not collected by assigning subjects or units into control and treatment groups beforehand, instead, they were simply measured or recorded. The main independent variable I am examining is the carbon dioxide production in a country, denoted as co2_prod_(year) in the data set. The independent variable is measured by the carbon dioxide output in a country on a standardized scale. The dependent variable is the Gross National Income per capita, measured in US dollars and denoted as gnipc_(year) in the data set. 

# Summary of the Dependent Variable
```{r}
#Packages and loading in data
library(tidyverse)
GNI_PerCapital <- read_csv("~/Downloads/Global Socio-Economic & Environmental Indicators/gross_national_income_per_capital.csv")
co2_production <- read_csv("~/Downloads/Global Socio-Economic & Environmental Indicators/co2_production.csv")

#Joining different data sets
co2_GNI <- co2_production |>
  inner_join(GNI_PerCapital)

#Making the joined data more usable for averaging gni and co2 production over years in a country.
co2_GNI <- co2_GNI |>
  pivot_longer(
    cols = co2_prod_1990:co2_prod_2021,
    names_to = "co2_peryear",
    values_to = "co2_prod"
  ) |>
  pivot_longer(
    cols = gnipc_1990:gnipc_2021,
    names_to = "gnipc_peryear",
    values_to = "gnipc"
  )

#Summary plot of dependent Variable
co2_GNI |>
  ggplot(aes(x = gnipc)) +
  geom_histogram(binwidth = 10000) +
  labs(title = "Distribution of GNI Per Capita",
       x = "GNI Per Capita")
```
The distribution of the dependent variable Gross National Income per capita is right-skewed, with the majority of states' GNI per capita falling between 0 and 25,000 US dollars. The peak is at the very left, hovering around 0. Given the clear outliers to the right, most states produce a modest amount of carbon dioxide, while a few produce an extreme amount relative to the others.

## Results
# Visualization
```{r}
#Scatterplot graph of relationship
co2_GNI |>
  na.omit() |>
  group_by(Country) |>
  summarize(
    avg_co2 = mean(co2_prod),
    avg_gnipc = mean(gnipc)
  ) |>
  ggplot(aes(x = avg_co2, y = avg_gnipc)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "How CO2 Production Correlates with GNI Per Capita",
    x = "CO2 Production",
    y = "Gross National Income Per Capita"
  )

```
This visualization demonstrates the relationship between GNI per capita and CO2 production. As Carbon dioxide production increases, the gross national income in a country appears to also generally increase. This graph demonstrates that countries with a higher gross national income also tend to have higher carbon dioxide production. Based on the graph, there appears to be a strong positive correlation between CO2 production and GNI per capita. The regression will give more information on this relationship.

# Main Independent Variable Regression
```{r}
#One-variable regression
reg <- lm(gnipc ~ co2_prod, data = co2_GNI)
modelsummary::modelsummary(reg,
                           statistic = c("stand. error = {std.error}",
                                         "p-value = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
This regression reveals the relationship between GNI Per Capita and CO2 Production. This regression shows that, on average, a one-unit increase in CO2 production is associated with an approximate **2173.086 increase** in gross national income per capita, in US dollars. This indicates a large positive correlation between the independent and dependent variables. The intercept of 6503.231, suggests that when there is no carbon dioxide production in the country, the estimated gross national income per capita is 6503.23 US dollars. In attempting to perform a significance test on this association, we can look at the p-value of the coefficient of co2_prod which is **less than 0.001**. At a significance level or alpha of 0.05, the result is **statistically significant** because it is less than 0.05. We can likely draw a causal relationship from this result; however, we can look at the effect of a confounder in the next part to attempt to prove that this relationship is not solely correlational. Further analysis will help prove that increased carbon dioxide production increases gross national income per capita.


# Multivariable Regression
```{r}
#Mult-variable regression
reg <- lm(gnipc ~ co2_prod + region, data = co2_GNI)
modelsummary::modelsummary(reg,
                           statistic = c("stand. error = {std.error}",
                                         "p-value = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
This multivariable regression demonstrates that even when accounting for a potential confounder, here it is region, there is still a large positive correlation between CO2 production and GNI per capita. In this regression, a one-unit increase in CO2 production correlates with an approximate **1739.34 US dollar increase** in GNI per capita. The p-value of this coefficient remains **less than 0.001**, which is **statistically significant** at the 0.05 level. There still appears to be a large positive correlation. There appears to be a statistically significant correlation between region and GNI per capita as well according to the multivariable regression, with most of them being large, with the notable exception of region AC. All of these region coefficients have a p-value of less than 0.05, so they are statistically significant at alpha = 0.05. This relationship can likely be due to countries in certain regions having similar development levels and carbon dioxide production rates. For example, most European countries will have similar development levels. However, it is still difficult to account for confounders, given the limited number of other variables available within the data set.

## Conclusion
This study aimed to evaluate the extent to which gross national income per capita and carbon dioxide production levels are related. Analyzing the relationship between these variables in the scatterplot and regressions offered conclusive results. There is a **large positive correlation** between gross national income per capita and carbon dioxide production across the globe. The results **support my hypothesis**. The statistical significance of this result was tested in the single-variable and multivariable regression. In both of these regressions, the coefficient for carbon dioxide production was **statistically significant** at an alpha of 0.05, and even at 0.01. We can reject a null hypothesis of there being no relationship. 

There were, however, a few limitations to the analysis. While the data gave data spanning many years, I decided to take the mean of that to get approximate results for each country. While it makes it easier to condense the data into overarching graphs and tables, it is an imperfect research method and makes the results less precise as opposed to having taken a long time to make specific calculations based on each country in each year. There was also missing data for years when the country had no available data or did not exist, meaning the data was incomplete. While the results are from every inhabited region of the world, there is a notable gap. The island nation of Nauru and the tiny enclave Vatican City, among other small states, are not included in this data set, making it difficult to apply the results and analysis to small nations. Having more holistic data and results would ameliorate these issues. If given more time/money, I would attempt to get data from every country in the world to mitigate any gaps in the data. Further, I would have taken more time to perform tests and analyses on individual years within specific countries to have more robust results. Further, focusing on specific years in specific countries may help to better contextualize the information. I would also collect data on and look more closely into potential confounders, such as market specialization and corruption. If given the time/money to perform these extra steps, there would be more conclusive evidence of causality globally and over the entire period.

