---
title: "Gov 50 Final Project"
author: "Sebastian Juarez"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Milestone 1: Project thoughts

I am interested in exploring data related to political elections and corruption in Southeast Asia, especially the Philippines. I am also interested in data about the black market in developing countries and socio-economic conditions that people in Southeast Asia face. Further, I would like to investigate data that highlights the economic performance of the various regimes that exist in Latin America.


## Milestone 2: Loading in data

```{r}
library(tidyverse)
GNI_PerCapital <- read_csv("~/Downloads/Global Socio-Economic & Environmental Indicators/gross_national_income_per_capital.csv")
co2_production <- read_csv("~/Downloads/Global Socio-Economic & Environmental Indicators/co2_production.csv")
```

Does a high gross national income per capita increase carbon dioxide output within a country? In my project, I will investigate the extent to which a growing gross national income per capita affects carbon dioxide output. I hypothesize that higher carbon dioxide output will correlate with a higher gross national income per capita. When the gross national income per capita is growing, I expect a higher number of goods and services exported from the country, which boosts the economy and typically raises wages. Additionally, as salaries rise, the consumption of carbon-emitting goods and services will increase. For example, more people would buy cars, use carbon-powered electricity, and take cars and airplanes. This increased economic activity would contribute to an overall higher carbon dioxide output. My sample consists of all of the countries in the world, with data collected every year from 1990 to 2021. My unit of analysis is a country. My explanatory variable of interest will be the gross national income per capita in a given year in a country, which will be in USD across all countries. My outcome variable is a country's carbon dioxide output for a given year. In the dataset, this is represented by a number that is proportional to a country’s carbon dioxide output. This number is used to make the data more manageable instead of using large numbers across different measuring systems. If I observe greater carbon dioxide production levels in years where the gross national income per capita has increased, this would support my hypothesis. If I observe lower carbon dioxide production levels in years where the gross national income per capita has increased, this would prove against my hypothesis. If I find a strong, positive correlation between the gross national income per capita and the carbon dioxide output within countries, this would support my hypothesis. 

## Milestone 3: Data Visualization

```{r}
co2_GNI <- co2_production |>
  inner_join(GNI_PerCapital)

co2_GNI <- co2_GNI |>
  pivot_longer(
    cols = co2_prod_1990:co2_prod_2021,
    names_to = "co2_peryear",
    values_to = "co2_prod"
  ) |>
  pivot_longer(
    cols = gnipc_1990:gnipc_2021,
    names_to = "gnipc_peryear",
    values_to = "gnipc"
  )
co2_GNI

co2_GNI |>
  na.omit() |>
  group_by(Country) |>
  summarize(
    avg_co2 = mean(co2_prod),
    avg_gnipc = mean(gnipc)
  ) |>
  ggplot(aes(x = avg_co2, y = avg_gnipc)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "GNI Per Capita as a Function of CO2 Production",
    x = "Carbon Dioxide Production",
    y = "Gross National Income"
  )
  

```
As Carbon dioxide production increased, the gross national income in a country also generally increased. This graph demonstrates that countries that have higher gross national incomes also tend to have higher rates of carbon dioxide production. While the majority of the points are clustered near the bottom left of the graph, even there is an even clearer relationship there.

```{r}
library(infer)
region <- co2_GNI |>
  na.omit() |>
  group_by(region, co2_peryear, gnipc_peryear) |>
  summarize(avg_co2_prod = mean(co2_prod),
            avg_gnipc = mean(gnipc))

region_co2 <- region |>
  ggplot(aes(x = co2_peryear, y = avg_co2_prod)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ region) +
  labs(
    title= "Regional Carbon Dioxide Output per year",
    x = "Year",
    y = "Carbon Dioxide Output"
  )

region_co2


region_GNI <- region |> 
  ggplot(aes(x = gnipc_peryear, y = avg_gnipc)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ region) +
  labs(
    title= "Regional GNI per year",
    x = "Year",
    y = "GNI"
  )
  
region_GNI

lm(gnipc ~ co2_prod, data = co2_GNI)
```
With this pair of visualizations, it is immediately visible that the rise and fall of GNI and CO2 are directly related. By analyzing the regression between the two, we can see that the relationship between the two is that there is a base carbon emission of 6503, but as gnipc increases by 1, carbon dioxide increases by 2173
